---
title: "Plastic Waste Project"
format:
  dashboard: 
    theme: cosmo
    logo: "https://encrypted-tbn0.gstatic.com/images?q=tbn:ANd9GcQTZuRREjunyFKk2gbvmSMWHEYi_KaBkjgR4w&s"
    nav-buttons:
      - icon: linkedin
        href: https://linkedin.com
        aria-label: LinkedIn
      - icon: instagram
        href: https://instagram.com
        aria-label: Instagram
      - icon: twitter-x
        href: https://twitter.com
        aria-label: Twitter/X
embed-resources: true
---

```{r}
#| label: package-setup
#| message: false
#| warning: false

library(leaflet)
library(gifski)
library(gganimate)
library(scales)
library(httr)
library(tidyjson)
library(jsonlite)
library(tidyverse)
library(gt)
library(glue)
library(purrr)
library(tidyr)
library(readr)
library(hexbin)
library(ggplot2)
```

```{r}
#| label: load-data
#| message: false
#| warning: false

# Load plastics
plastics <- read.csv("~/Documents/R/plastics.csv") 

# Load country capitals
country_data <- read_csv("concap.csv")

capitals <- country_data |>
  select(c(1:4)) |>
  rename(
    "country" = "CountryName",
    "capital_city" = "CapitalName",
    "capital_lat" = "CapitalLatitude",
    "capital_long" = "CapitalLongitude"
  )

# Join plastics with capitals and air data
pollution_data <- left_join(plastics, capitals, by = "country")
air_data       <- read_csv("air_data.csv")
pollution_data <- left_join(pollution_data, air_data, by = "country")

# Build air vs plastic summary 
air_vs_plastic <- pollution_data |>
  filter(!parent_company %in% c("Grand Total", "Unbranded", "null", "NULL")) |>
  group_by(country) |>
  summarize(
    total_plastic    = sum(grand_total, na.rm = TRUE),
    total_volunteers = sum(volunteers, na.rm = TRUE),
    avg_aqi          = mean(overall_aqi, na.rm = TRUE),
    capital_city     = first(capital_city),
    capital_lat      = first(capital_lat),
    capital_long     = first(capital_long)
  ) |>
  filter(!is.na(avg_aqi), total_plastic > 0) |>
  arrange(desc(total_plastic))

# IQR filter
Q1  <- quantile(air_vs_plastic$total_plastic, 0.25)
Q3  <- quantile(air_vs_plastic$total_plastic, 0.75)
IQR <- Q3 - Q1

air_vs_plastic_clean <- air_vs_plastic |>
  filter(
    total_plastic >= Q1 - 1.5 * IQR,
    total_plastic <= Q3 + 1.5 * IQR
  )

# Map-ready data (needs lat/long)
map_data <- air_vs_plastic |>
  filter(
    !is.na(capital_lat),
    !is.na(capital_long),
    !is.na(total_plastic),
    !is.na(avg_aqi),
    total_plastic >= Q1 - 1.5 * IQR,
    total_plastic <= Q3 + 1.5 * IQR
  )

# Philippines fertility data
fertility           <- read_csv("fertility_rate.csv", skip = 4)
fertility_ph        <- fertility |>
  filter(`Country Name` == "Philippines") |>
  rename("country" = "Country Name")

combined <- right_join(air_vs_plastic, fertility_ph, by = "country") |>
  filter(country == "Philippines") |>
  pivot_longer(
    cols      = starts_with("19") | starts_with("20"),
    names_to  = "year",
    values_to = "fertility_rate"
  ) |>
  mutate(year = as.integer(year)) |>
  filter(!is.na(fertility_rate))
```

# Overview

## Column {width="30%"}

### Introduction

**Research Question: Is plastic waste associated with air quality across countries?**

The data comes from **Break Free From Plastic** global brand audits conducted in 2019 and 2020, where volunteers physically collected and recorded plastic pieces by type and company label. The dataset was put together by Sarah Sauve and covers dozens of countries worldwide.
You can find it [here](https://github.com/sarahsauve/TidyTuesdays/blob/master/BFFPDashboard/BlogPost.md).

We combined this data with:

- **Country Capitals GPS data** — to map each country's coordinates
- **Air Quality API (API Ninjas)** — to pull AQI scores by capital city

The observational unit is one brand audited in one country in one year. Key variables include plastic type counts, grand total, number of volunteers, and average AQI score per country.

**Why does this matter?** Countries with high plastic pollution may face compounding environmental pressures. Understanding whether plastic waste and air quality are linked can help policymakers prioritize environmental interventions.


**Key Findings:**

- The **Philippines** has the highest plastic waste of any country in our dataset
- There is **no strong linear relationship** between plastic waste and air quality across countries


## Column {width="70%"}

### Global Plastic Waste and Air Quality 

```{r}
#| label: leaflet-map

aqi_palette <- colorNumeric(
  palette = c("green", "yellow", "orange", "red"),
  domain  = map_data$avg_aqi
)

leaflet(map_data) |>
  addTiles() |>
  addCircleMarkers(
    lng         = ~capital_long,
    lat         = ~capital_lat,
    radius      = ~rescale(total_plastic, to = c(3, 15)),
    fillColor   = ~aqi_palette(avg_aqi),
    fillOpacity = 0.7,
    color       = "white",
    weight      = 1,
    popup       = ~paste0(
      "<b>", country, "</b><br>",
      "Capital: ", capital_city, "<br>",
      "Total Plastic Waste: ", round(total_plastic, 0), "<br>",
      "Avg AQI: ", round(avg_aqi, 1)
    ),
    label = ~country
  ) |>
  addLegend(
    position = "bottomright",
    pal      = aqi_palette,
    values   = ~avg_aqi,
    title    = "Avg AQI",
    opacity  = 0.7
  )
```
Circle size = total plastic waste. Color = average AQI (green = good, red = unhealthy). Click a country for details.

### Key Statistics

```{r}
tibble(
  Metric = c("Countries in Dataset", "Total Plastic Pieces", 
             "Countries with AQI Data", "Years Covered"),
  Value  = c(
    n_distinct(plastics$country),
    scales::comma(sum(plastics$grand_total, na.rm = TRUE)),
    nrow(air_vs_plastic),
    "2019 – 2020"
  )
) |>
  gt() |>
  cols_label(Metric = "", Value = "") |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_body(columns = Metric)
  )
```

# Air Quality & Plastic

## Column {width="55%"}

### Plastic Waste and Air Quality Across Countries

```{r}
#| label: bubble-plot

air_vs_plastic_clean |>
  mutate(aqi_category = case_when(
    avg_aqi <= 50  ~ "Good",
    avg_aqi <= 100 ~ "Moderate",
    avg_aqi <= 150 ~ "Unhealthy for Sensitive Groups",
    TRUE           ~ "Unhealthy"
  )) |>
  ggplot(aes(x = total_plastic, y = avg_aqi,
             size = total_plastic, color = aqi_category)) +
  geom_point(alpha = 0.7) +
  ggrepel::geom_text_repel(
    aes(label = country),
    size = 2.5,
    max.overlaps = 10
  ) +
  scale_size_continuous(range = c(2, 12)) +
  scale_color_manual(values = c(
    "Good"                           = "darkgreen",
    "Moderate"                       = "gold",
    "Unhealthy for Sensitive Groups" = "orange",
    "Unhealthy"                      = "red"
  )) +
  labs(
    title    = "Plastic Waste and Air Quality Across Countries",
    subtitle = "Bubble size reflects total plastic waste. Color reflects AQI category.",
    x        = "Total Plastic Waste",
    y        = "Average AQI",
    color    = "AQI Category"
  ) +
  theme_minimal() +
  theme(
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(size = 11),
    panel.border     = element_rect(color = "black", fill = NA, linewidth = 0.8),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  ) +
  guides(size = "none")
```


### Association between Plastic Waste and Air Quality

```{r}
#| label: scatter-plot

ggplot(
  data = air_vs_plastic_clean,
  aes(x = total_plastic, y = avg_aqi)
) +
  geom_point(size = 3, alpha = 0.7, color = "darkgreen") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(
    title    = "Association between Plastic Waste and Air Quality",
    subtitle = "Each point represents one country. Red line shows the linear trend.",
    x        = "Total Plastic Waste",
    y        = "Average Air Quality Index (AQI)"
  ) +
  theme_minimal() +
  theme(
    plot.title    = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray90"),
    panel.border  = element_rect(color = "black", fill = NA, linewidth = 0.8)
  )
```

## Column {width="45%"}

### Air Quality Analysis

::: {.panel-tabset}

## Top 10 Countries

```{r}
air_vs_plastic |>
  slice_max(total_plastic, n = 10) |>
  select(country, total_plastic, avg_aqi) |>
  gt() |>
  tab_header(
    title    = "Top 10 Countries by Plastic Waste",
    subtitle = "With corresponding average AQI"
  ) |>
  cols_label(
    country       = "Country",
    total_plastic = "Total Plastic",
    avg_aqi       = "Avg AQI"
  ) |>
  fmt_number(columns = total_plastic, decimals = 0) |>
  fmt_number(columns = avg_aqi,       decimals = 1) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  )
```

## AQI by Plastic Level

```{r}
air_vs_plastic_clean |>
  mutate(
    plastic_quartile = case_when(
      total_plastic <= quantile(total_plastic, 0.25, na.rm = TRUE) ~ "Low (Q1)",
      total_plastic <= quantile(total_plastic, 0.50, na.rm = TRUE) ~ "Medium-Low (Q2)",
      total_plastic <= quantile(total_plastic, 0.75, na.rm = TRUE) ~ "Medium-High (Q3)",
      TRUE ~ "High (Q4)"
    ),
    plastic_quartile = factor(
      plastic_quartile,
      levels = c("Low (Q1)", "Medium-Low (Q2)", "Medium-High (Q3)", "High (Q4)")
    )
  ) |>
  group_by(plastic_quartile) |>
  summarise(
    n_countries = n(),
    average_aqi = round(mean(avg_aqi, na.rm = TRUE), 2),
    min_aqi = round(min(avg_aqi, na.rm = TRUE), 2),
    max_aqi = round(max(avg_aqi, na.rm = TRUE), 2),
    .groups = "drop"
  ) |>
  arrange(plastic_quartile) |>
  gt() |>
  opt_table_font(font = "Arial") |>
  tab_header(
    title = "Air Quality by Plastic Waste Level",
    subtitle = "Countries grouped into quartiles by total plastic waste"
  ) |>
  cols_label(
    plastic_quartile = "Plastic Waste Group",
    n_countries = "Number of Countries",
    average_aqi = "Average AQI",
    min_aqi = "Minimum AQI",
    max_aqi = "Maximum AQI"
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  data_color(
    columns = average_aqi,
    colors  = scales::col_numeric(
      palette = c("darkgreen", "yellow", "red"),
      domain  = NULL
    )
  ) |>
  tab_source_note("Note: Higher AQI indicates worse air quality.Countries filtered to remove outliers.")
```

## Model Results

```{r}
model         <- lm(avg_aqi ~ total_plastic, data = air_vs_plastic)
summary_model <- summary(model)

r_squared     <- round(summary_model$r.squared, 4)
adj_r_squared <- round(summary_model$adj.r.squared, 4)
f_stat        <- round(summary_model$fstatistic[1], 3)
p_value_model <- round(
  pf(summary_model$fstatistic[1],
     summary_model$fstatistic[2],
     summary_model$fstatistic[3],
     lower.tail = FALSE), 4)

data.frame(
  Term           = c("Intercept", "Total Plastic"),
  Estimate       = coef(model),
  Standard_Error = summary_model$coefficients[, 2],
  T_Statistic    = summary_model$coefficients[, 3],
  P_Value        = summary_model$coefficients[, 4]
) |>
  gt() |>
  tab_header(
    title    = "Linear Regression Results",
    subtitle = "Predicting Average AQI from Total Plastic Waste"
  ) |>
  tab_spanner(label = "Coefficient", columns = c(Estimate, Standard_Error)) |>
  tab_spanner(label = "Test",        columns = c(T_Statistic, P_Value)) |>
  cols_label(
    Term           = "Term",
    Estimate       = "Estimate",
    Standard_Error = "Std. Error",
    T_Statistic    = "t",
    P_Value        = "p-value"
  ) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  fmt_number(columns = c(Estimate, Standard_Error, T_Statistic), decimals = 3) |>
  fmt_number(columns = P_Value, decimals = 4) |>
  tab_source_note(glue(
    "R² = {r_squared} | Adj. R² = {adj_r_squared} | F = {f_stat} | p = {p_value_model}"
  ))
```

:::

# Philippines

## Column {width="40%"}

### Why the Philippines?

The **Philippines** stands out in our dataset as a clear outlier. It has both the **highest total plastic waste** and one of the **highest average AQI scores**, meaning it faces compounding environmental pressures that few other countries in our data match.

It is a rapidly developing country where population growth drives increasing consumption and waste generation. To explore this connection further, we incorporated **fertility rate data from the World Bank**, which shows how the Philippines' birth rate has changed over time and what that implies for future plastic waste and air quality trends.

### Philippines at a Glance

```{r}
air_vs_plastic |>
  filter(country == "Philippines") |>
  select(country, total_plastic, avg_aqi) |>
  gt() |>
  tab_header(title = "Philippines — Key Stats") |>
  cols_label(
    country       = "Country",
    total_plastic = "Total Plastic Waste",
    avg_aqi       = "Average AQI"
  ) |>
  fmt_number(columns = total_plastic, decimals = 0) |>
  fmt_number(columns = avg_aqi,       decimals = 1) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_source_note("Source: BFFP brand audit data (2019–2020) + API Ninjas AQI")
```

## Column {width="60%"}

### Fertility Rate in the Philippines Over Time

*An animated look at how the Philippines' birth rate has declined and what it means for future plastic waste.*

```{r}
#| label: animated-plot

p <- ggplot(combined, aes(x = year, y = fertility_rate, group = 1)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "red", size = 3) +
  geom_hline(
    yintercept = 2.1,
    color      = "red",
    linetype   = "dashed"
  ) +
  annotate(
    "text",
    x     = min(combined$year),
    y     = 2.1,
    label = "Replacement rate (2.1)",
    color = "red",
    vjust = -0.5,
    hjust = 0,
    size  = 3.5
  ) +
  labs(
    title    = "Fertility Rate in the Philippines Over Time",
    subtitle = "Year: {frame_along}",
    x        = "Year",
    y        = "Births per Woman"
  ) +
  theme_minimal() +
  theme(
    plot.title    = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11),
    panel.border  = element_rect(color = "black", fill = NA, linewidth = 0.8)
  ) +
  transition_reveal(year)

anim <- animate(p, duration = 10, fps = 20, renderer = gifski_renderer())
anim
```